---
title: "2025 National Taiwan University - Population Pharmacokinetics workshop"
output: 
  html_document:
    css: docs/src/styles/styles.css
    toc: true
    number_sections: true
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(comment='.', echo = TRUE, warning = FALSE)
rm(list=ls())
```

# Load Packages
```{r echo=TRUE}
library(renv)
library(tidyverse)
library(ggplot2)
library(nlmixr2)
library(xpose4)
library(xpose.nlmixr2)
library(rxode2)
library(gridExtra)
library(ggPMX)
library(ggpubr)
```

# Data exploration
```{r}
# Import busulfan two compartment dataset
busulfan_dataset <- read.csv("dataset/busulfan_Test_Dose.csv", na.strings = ".")

# Plasma drug concentration decline in two phases shown in semi-log plot
ggplot(busulfan_dataset, aes(TIME/60, log(DV), group = ID, color = as.factor(ID))) +
  geom_line() +
  theme_bw() +
  labs(title = "Busulfan semi-log plot", x = "Time (hours)", y = "Conc (mg/L)") +
  theme(legend.position = "blank")
```

# Showcase 1. 1-cmt vs 2-cmt

## Fitting with 1-cmt vs 2-cmt
```{r}
# One compartment model structure
busulfan_1cmt_base_model <- function() {
  ini({
    # Typical value (THETAs)
    tvcl   <- log(5)
    tvv1  <- log(50)
    
    # Interindividual variability (OMEGAs)
    eta_cl  ~ 0.0322
    eta_v1  ~ 0.0222
    
    # Residual variability
    add.error <-  0.0955
  })
  
  model({
    # Individual value
    cl <- exp(tvcl + eta_cl)
    v1 <- exp(tvv1 + eta_v1)

    # Conversion
    k10 <- cl / v1
    
    # ODEs
    d / dt(central) = -k10 * central
    
    # Concentration
    cp = central / v1
    
    # Error model
    IPRED = cp
    IPRED ~ add(add.error)
  })
}

# Two compartment model structure
busulfan_2cmt_base_model <- function() {
  ini({
    # Typical value (THETAs)
    tvcl   <- log(0.187)
    tvv1  <- log(29)
    tvq   <- log(0.41)
    tvv2  <- log(17.3)
    
    # Interindividual variability (OMEGAs)
    eta_cl  ~ 0.0222 
    eta_v1  ~ 0.0222
    eta_v2  ~ 0.0241
    
    # Residual variability
    prop.err <-  0.0955
  })
  
  model({
    # Individual parameters
    cl <- exp(tvcl + eta_cl) 
    v1 <- exp(tvv1 + eta_v1)
    q  <- exp(tvq)
    v2 <- exp(tvv2 + eta_v2)
    
    # Conversion
    k10 <- cl / v1
    k12 <- q / v1
    k21 <- q / v2
    
    # ODEs
    d / dt(central) = -k10 * central - k12 * central + k21 * peripheral
    d / dt(peripheral) = -k21 * peripheral + k12 * central
    
    # Concentration
    cp = central / v1
    
    # Error model
    IPRED = cp
    IPRED ~ prop(prop.err)
  })
}

# 1 compartment model fitting 
one_cmt_pk_fit <- nlmixr2(
  busulfan_1cmt_base_model,
  busulfan_dataset,
  "focei",
  table = list(cwres = TRUE)
)

# 2 compartment model fitting 
two_cmt_pk_fit <- nlmixr2(
  busulfan_2cmt_base_model,
  busulfan_dataset,
  "focei",
  table = list(cwres = TRUE)
)
```


## Model diagnostics - Objective Function Values
```{r}
# Extract Objective Functions
one_cmt_objDf <- one_cmt_pk_fit$objDf
two_cmt_objDf <- two_cmt_pk_fit$objDf


compare_obj <- rbind(one_cmt_objDf, two_cmt_objDf)
rownames(compare_obj) <- c("1-cmt", "2-cmt") 
compare_obj[,1:4]
```

## Model diagnostics - Goodness of Fit (GOF)
```{r}
# Convert to XPOSE for goodness of fit
one_cmt_xpdb_pk <- xpose_data_nlmixr2(one_cmt_pk_fit,
                                     xp_theme = theme_xp_nlmixr2()
                                     )

two_cmt_xpdb_pk <- xpose_data_nlmixr2(two_cmt_pk_fit,
                                     xp_theme = theme_xp_nlmixr2()
                                     )

# observation vs population prediction plot
one_cmt_dv_pred <- dv_vs_pred(one_cmt_xpdb_pk, 
                              type="pls", 
                              title = "One Compartment",
                              log   = 'xy',
                              subtitle = NULL,
                              guide = TRUE)
two_cmt_dv_pred <- dv_vs_pred(two_cmt_xpdb_pk, 
                              type="pls", 
                              title = "Two Compartment",
                              log   = 'xy',
                              subtitle = NULL,
                              guide = TRUE)

grid.arrange(one_cmt_dv_pred, two_cmt_dv_pred, nrow = 1)

# observation vs individual prediction plot
one_cmt_dv_ipred <- dv_vs_ipred(one_cmt_xpdb_pk, 
                                group="ID",
                                title = "One Compartment",
                                log   = 'xy',
                                subtitle = NULL,
                                guide = TRUE
                                )
two_cmt_dv_ipred <- dv_vs_ipred(two_cmt_xpdb_pk, 
                                group="ID",
                                title = "Two Compartment",
                                log   = 'xy',
                                subtitle = NULL,
                                guide = TRUE
                                )

grid.arrange(one_cmt_dv_ipred, two_cmt_dv_ipred, nrow = 1)

# IPRED vs TIME
one_cmt_ipred_time <- pred_vs_idv(one_cmt_xpdb_pk, 
                                 group="ID",
                                 title = "One Compartment",
                                 subtitle = NULL,
                                 guide = TRUE
                                 )

two_cmt_ipred_time <- pred_vs_idv(two_cmt_xpdb_pk, 
                                 group="ID",
                                 title = "Two Compartment",
                                 subtitle = NULL,
                                 guide = TRUE
                                 )

grid.arrange(one_cmt_ipred_time, two_cmt_ipred_time, nrow = 1)

# residual error vs population prediction plot
one_cmt_cwres_pred <- res_vs_pred(one_cmt_xpdb_pk, 
                                  res="CWRES", 
                                  type="ps",
                                  title = "One Compartment",
                                  subtitle = NULL,
                                  guide = TRUE
                                  )

two_cmt_cwres_pred <- res_vs_pred(two_cmt_xpdb_pk, 
                                  res="CWRES", 
                                  type="ps",
                                  title = "Two Compartment",
                                  subtitle = NULL,
                                  guide = TRUE
                                  )

grid.arrange(one_cmt_cwres_pred, two_cmt_cwres_pred, nrow = 1)
```

## Two-cmt model parameters
```{r}
# Parameter estimates - Fixed + RUV
two_cmt_pk_fit$parFixedDf %>% select(`Back-transformed`, `%RSE`, `Shrink(SD)%`) %>% rename(Parameters = `Back-transformed`)

# IIV
two_cmt_pk_fit$omega
```


# Showcase 2. Error model (Additive vs Proporational vs Combined) 

## Fitting with proportional vs combined vs additive error
```{r}
# Two compartment model with additive model
busulfan_2cmt_add_model <- function() {
  ini({
    # Typical value (THETAs)
    tvcl   <- log(0.187)
    tvv1  <- log(29)
    tvq   <- log(0.41)
    tvv2  <- log(17.3)
    
    # Interindividual variability (OMEGAs)
    eta_cl  ~ 0.0222 
    eta_v1  ~ 0.0222
    eta_v2  ~ 0.0241
    
    # Residual variability
    add.err <-  0.0955
  })
  
  model({
    # Individual parameters
    cl <- exp(tvcl + eta_cl) 
    v1 <- exp(tvv1 + eta_v1)
    q  <- exp(tvq)
    v2 <- exp(tvv2 + eta_v2)
    
    # Conversion
    k10 <- cl / v1
    k12 <- q / v1
    k21 <- q / v2
    
    # ODEs
    d / dt(central) = -k10 * central - k12 * central + k21 * peripheral
    d / dt(peripheral) = -k21 * peripheral + k12 * central
    
    # Concentration
    cp = central / v1
    
    # Error model (on log scale)
    IPRED = cp
    IPRED ~ add(add.err)
  })
}

# additive model fitting 
two_cmt_add_pk_fit <- nlmixr2(
  busulfan_2cmt_add_model,
  busulfan_dataset,
  "focei",
  table = list(cwres = TRUE)
)

# Two compartment model with combined
busulfan_2cmt_combined_model <- function() {
  ini({
    # Typical value (THETAs)
    tvcl   <- log(0.187)
    tvv1  <- log(29)
    tvq   <- log(0.41)
    tvv2  <- log(17.3)
    
    # Interindividual variability (OMEGAs)
    eta_cl  ~ 0.0222 
    eta_v1  ~ 0.0222
    eta_v2  ~ 0.0241
    
    # Residual variability
    prop.err <-  0.0955
    add.err  <-  0.02
  })
  
  model({
    # Individual parameters
    cl <- exp(tvcl + eta_cl) 
    v1 <- exp(tvv1 + eta_v1)
    q  <- exp(tvq)
    v2 <- exp(tvv2 + eta_v2)
    
    # Conversion
    k10 <- cl / v1
    k12 <- q / v1
    k21 <- q / v2
    
    # ODEs
    d / dt(central) = -k10 * central - k12 * central + k21 * peripheral
    d / dt(peripheral) = -k21 * peripheral + k12 * central
    
    # Concentration
    cp = central / v1
    
    # Error model (on log scale)
    IPRED = cp
    IPRED ~ prop(prop.err) + add(add.err)
  })
}

# combined model fitting 
two_cmt_combined_pk_fit <- nlmixr2(
  busulfan_2cmt_combined_model,
  busulfan_dataset,
  "focei",
  table = list(cwres = TRUE)
  )
```

## Model diagnostics - Objective Function Values
```{r}
# Extract Objective Functions
two_cmt_add_objDf <- two_cmt_add_pk_fit$objDf 
two_cmt_prop_objDf  <- two_cmt_pk_fit$objDf
two_cmt_comb_objDf <- two_cmt_combined_pk_fit$objDf

error_compare_obj <- rbind(two_cmt_add_objDf, two_cmt_prop_objDf, two_cmt_comb_objDf)
rownames(error_compare_obj) <- c("Additive", "Proportional", "Combined") 
error_compare_obj
```



```{r}
# Convert to XPOSE for goodness of fit

two_cmt_add_xpdb_pk <- xpose_data_nlmixr2(two_cmt_add_pk_fit,
                                     xp_theme = theme_xp_nlmixr2()
                                     )

two_cmt_combined_xpdb_pk <- xpose_data_nlmixr2(two_cmt_combined_pk_fit,
                                     xp_theme = theme_xp_nlmixr2()
                                     )

# observation vs individual prediction plot
add_error_dv_pred <- dv_vs_pred(two_cmt_add_xpdb_pk, 
                                group="ID",
                                title = "Additive Error",
                                #log = "xy",
                                subtitle = NULL,
                                guide = FALSE
                                )

prop_error_dv_pred <- dv_vs_pred(two_cmt_xpdb_pk, 
                                group="ID",
                                title = "Proportional Error",
                                #log = "xy",
                                subtitle = NULL,
                                guide = FALSE
                                )
comb_error_dv_pred <- dv_vs_pred(two_cmt_combined_xpdb_pk, 
                                group="ID",
                                title = "Combined Error",
                                #log = "xy",
                                subtitle = NULL,
                                guide = FALSE
                                )

grid.arrange(add_error_dv_pred, prop_error_dv_pred, comb_error_dv_pred, nrow = 1)

# residual error vs population prediction plot
add_error_cwres <- res_vs_pred(two_cmt_add_xpdb_pk, 
                                  res="CWRES", 
                                  type="ps",
                                  title = "Additive error",
                                  subtitle = NULL,
                                  guide = TRUE
                                  )

prop_error_cwres <- res_vs_pred(two_cmt_xpdb_pk, 
                                  res="CWRES", 
                                  type="ps",
                                  title = "Proportioanl error",
                                  subtitle = NULL,
                                  guide = TRUE
                                  )


comb_error_cwres <- res_vs_pred(two_cmt_combined_xpdb_pk, 
                                  res="CWRES", 
                                  type="ps",
                                  title = "Combined error",
                                  subtitle = NULL,
                                  guide = TRUE
                                  )

grid.arrange(add_error_cwres, prop_error_cwres, comb_error_cwres, nrow = 1)
```

## Two-cmt model parameters
```{r}
# Parameter estimates - Fixed + RUV
two_cmt_pk_fit$parFixedDf %>% select(`Back-transformed`, `%RSE`, `Shrink(SD)%`) %>% rename(Parameters = `Back-transformed`)

# Parameter estimates - Fixed + RUV
two_cmt_combined_pk_fit$parFixedDf %>% select(`Back-transformed`, `%RSE`, `Shrink(SD)%`) %>% rename(Parameters = `Back-transformed`)

```

# Showcase 3. Covariates to explain Inter-Individual Variability

## Omegas and Etas - Distribution of individual parameters
```{r}
# Parameter estimates - Fixed + RUV
two_cmt_pk_fit$parFixedDf %>% select(`Back-transformed`, `%RSE`, `Shrink(SD)%`) %>% rename(Parameters = `Back-transformed`)

base_combined_dataset <- merge(two_cmt_pk_fit, busulfan_dataset, by = c("ID", "TIME", "DV"))

# Sex as covariates
cov_sex_cl <- ggplot(base_combined_dataset, aes(as.factor(Sex), eta_cl)) + 
  geom_boxplot() +
  stat_compare_means(method = "wilcox.test", label = "p.format") +
  theme_bw() +
  labs(title = "CL vs Sex", x = "Sex")

cov_sex_v1 <- ggplot(base_combined_dataset, aes(as.factor(Sex), eta_v1)) + 
  geom_boxplot() +
  stat_compare_means(method = "wilcox.test", label = "p.format") +
  theme_bw() +
  labs(title = "V1 vs Sex", x = "Sex")

cov_sex_v2 <- ggplot(base_combined_dataset, aes(as.factor(Sex), eta_v2)) + 
  geom_boxplot() +
  stat_compare_means(method = "wilcox.test", label = "p.format") +
  theme_bw() +
  labs(title = "V2 vs Sex", x = "Sex")

grid.arrange(cov_sex_cl, cov_sex_v1, cov_sex_v2, nrow = 1)

# BSA as covariates
cov_bsa_cl <- ggplot(base_combined_dataset, aes(BSA, eta_cl)) + 
  geom_point() +
  geom_smooth(method = "lm", se = TRUE) +
  stat_cor(method = "pearson", 
           aes(label = paste(..rr.label.., ..p.label.., sep = "~`,`~"))) +
  theme_bw() +
  labs(title = "CL vs BSA", x = "BSA")

cov_bsa_v1 <- ggplot(base_combined_dataset, aes(BSA, eta_v1)) + 
  geom_point() +
  geom_smooth(method = "lm", se = TRUE) +
  stat_cor(method = "pearson", 
           aes(label = paste(..rr.label.., ..p.label.., sep = "~`,`~"))) +
  theme_bw() +
  labs(title = "V1 vs BSA", x = "BSA")

cov_bsa_v2 <- ggplot(base_combined_dataset, aes(BSA, eta_v2)) + 
  geom_point() +
  geom_smooth(method = "lm", se = TRUE) +
  stat_cor(method = "pearson", 
           aes(label = paste(..rr.label.., ..p.label.., sep = "~`,`~"))) +
  theme_bw() +
  labs(title = "V2 vs BSA", x = "BSA")

grid.arrange(cov_bsa_cl, cov_bsa_v1, cov_bsa_v2, nrow = 1)

```


## Run covariate model
```{r}
busulfan_final_2cmt_model <- function() {
  ini({
    # Typical value (THETAs)
    tvcl   <- log(0.187)
    tvv1  <- log(29)
    tvq   <- log(0.41)
    tvv2  <- log(17.3)
    
    covbsav1 <- log(2.32)
    covbsacl <- log(1.30)
    covsexv2 <- log(0.8)
    
    # Interindividual variability (OMEGAs)
    eta_cl  ~ 0.0222 
    eta_v1  ~ 0.0222
    eta_v2  ~ 0.0241
    
    # Residual variability
    prop.err <-  0.0955
  })
  
  model({
    # Individual value
    cl <- exp(tvcl + eta_cl) * (BSA/2.01)^covbsacl
    v1 <- exp(tvv1 + eta_v1) * (BSA/2.01)^covbsav1
    q  <- exp(tvq)
    v2 <- exp(tvv2 + eta_v2)
    
    # Sex effect
    if (Sex == 1) {
      v2 <- v2 * covsexv2
    } else {
      v2 <- v2 * 1
    }
    
    # Conversion
    k10 <- cl / v1
    k12 <- q / v1
    k21 <- q / v2
    
    # ODEs
    d / dt(central) = -k10 * central - k12 * central + k21 * peripheral
    d / dt(peripheral) = -k21 * peripheral + k12 * central
    
    # Concentration
    cp = central / v1
    
    # Error model (on log scale)
    IPRED = cp
    IPRED ~ prop(prop.err)
  })
}

# Model Fitting
two_cmt_final_pk_fit <- nlmixr2(
  busulfan_final_2cmt_model,
  busulfan_dataset,
  "focei",
  table = list(cwres = TRUE, npde = TRUE)
)
```

```{r}
final_combined_dataset <- merge(two_cmt_final_pk_fit, busulfan_dataset, by = c("ID", "TIME", "DV", "BSA", "Sex"))

# Sex as covariates
cov_sex_cl <- ggplot(final_combined_dataset, aes(as.factor(Sex), eta_cl)) + 
  geom_boxplot() +
  stat_compare_means(method = "wilcox.test", label = "p.format") +
  theme_bw() +
  labs(title = "CL vs Sex", x = "Sex")

cov_sex_v1 <- ggplot(final_combined_dataset, aes(as.factor(Sex), eta_v1)) + 
  geom_boxplot() +
  stat_compare_means(method = "wilcox.test", label = "p.format") +
  theme_bw() +
  labs(title = "V1 vs Sex", x = "Sex")

cov_sex_v2 <- ggplot(final_combined_dataset, aes(as.factor(Sex), eta_v2)) + 
  geom_boxplot() +
  stat_compare_means(method = "wilcox.test", label = "p.format") +
  theme_bw() +
  labs(title = "V2 vs Sex", x = "Sex")

grid.arrange(cov_sex_cl, cov_sex_v1, cov_sex_v2, nrow = 1)

# BSA as covariates
cov_bsa_cl <- ggplot(final_combined_dataset, aes(BSA, eta_cl)) + 
  geom_point() +
  geom_smooth(method = "lm", se = TRUE) +
  stat_cor(method = "pearson", 
           aes(label = paste(..rr.label.., ..p.label.., sep = "~`,`~"))) +
  theme_bw() +
  labs(title = "CL vs BSA", x = "BSA")

cov_bsa_v1 <- ggplot(final_combined_dataset, aes(BSA, eta_v1)) + 
  geom_point() +
  geom_smooth(method = "lm", se = TRUE) +
  stat_cor(method = "pearson", 
           aes(label = paste(..rr.label.., ..p.label.., sep = "~`,`~"))) +
  theme_bw() +
  labs(title = "V1 vs BSA", x = "BSA")

cov_bsa_v2 <- ggplot(final_combined_dataset, aes(BSA, eta_v2)) + 
  geom_point() +
  geom_smooth(method = "lm", se = TRUE) +
  stat_cor(method = "pearson", 
           aes(label = paste(..rr.label.., ..p.label.., sep = "~`,`~"))) +
  theme_bw() +
  labs(title = "V2 vs BSA", x = "BSA")

grid.arrange(cov_bsa_cl, cov_bsa_v1, cov_bsa_v2, nrow = 1)
```

```{r}
## Two-cmt model parameters
# Parameter estimates - Fixed + RUV

# Compare etas between base vs final
print("Base Model etas")
two_cmt_pk_fit$omega

print("Final Model etas")
two_cmt_final_pk_fit$omega

# Final model parameters
two_cmt_final_pk_fit$parFixedDf %>% select(`Back-transformed`, `%RSE`, `Shrink(SD)%`) %>% rename(Parameters = `Back-transformed`)
```
```

