---
title: "2025 National Taiwan University - Population Pharmacokinetics workshop"
output: 
  html_document:
    css: docs/src/styles/styles.css
    toc: true
    number_sections: true
---
# Load Packages
Here we will import our installed packages into our R environment
```{r echo=TRUE}
library(renv)
library(tidyverse)
library(ggplot2)
library(nlmixr2)
library(xpose4)
library(xpose.nlmixr2)
library(rxode2)
library(gridExtra)
library(ggPMX)
library(ggpubr)
library(mrgsolve)
library(vpc)
library(patchwork)
```

# Dataset
```{r}
# Import busulfan two compartment dataset
busulfan_First_Dose_dataset <- read.csv("dataset/busulfan_First_Dose.csv", na.strings = ".")

busulfan_TDM_dataset <- busulfan_First_Dose_dataset %>% 
  group_by(ID) %>%
  filter(row_number() == 1 | row_number() == 2 | row_number() == n())
```

# Showcase 4
## Overlay First Dose data with the built model - Individual fits
```{r}
model_code <- 'model/busulfan_test_dose_model.mod'
mod <- mread('busulfan_test_dose', file = model_code)

sim <- function(rep, data, model,
                recover = c('AMT', 'DV', 'Sex', 'BSA')) {
  mrgsim(
    model, 
    data = data,
    recover = recover,
    Req = c('CP_no_RUV', 'CP', 'no_obs_AUC'),
    output = 'df',
  ) %>% mutate(irep = rep)
}

isim <- seq(1000)

set.seed(86486)

sims <- lapply(
  isim, sim, 
  data = busulfan_First_Dose_dataset, 
  mod = mod
) %>% bind_rows()

indv_fits <- sims %>% filter(irep == 1)

p1 <- ggplot() +
  geom_line(data = indv_fits, aes(TIME/60, CP_no_RUV, group = ID)) +
  geom_point(data = busulfan_TDM_dataset, aes(TIME/60, DV, group = ID)) +
  facet_wrap_paginate(~ID, nrow = 4, ncol = 4, page = 1) +
  xlim(2.5, 10) +
  labs(x = 'Time (horus)',  y = "Busulfan Concentration (mg/L)") +
  theme_bw()

p2 <- ggplot() +
  geom_line(data = indv_fits, aes(TIME/60, CP_no_RUV, group = ID)) +
  geom_point(data = busulfan_TDM_dataset, aes(TIME/60, DV, group = ID)) +
  facet_wrap_paginate(~ID, nrow = 4, ncol = 4, page = 2) +
  xlim(2.5, 10) +
  labs(x = 'Time (horus)',  y = "Busulfan Concentration (mg/L)") +
  theme_bw()

p1
p2
```

## Overlay First Dose data with the built model - pcVPC
```{r}
mrg_vpc_theme = new_vpc_theme(list(
  sim_pi_fill = "steelblue3", sim_pi_alpha = 0.5,
  sim_median_fill = "grey60", sim_median_alpha = 0.5
))

vpc_plot <- vpc(obs = busulfan_TDM_dataset,
          obs_cols = list(dv = 'DV',
                          idv = 'TIME',
                          id = 'ID'
                          ),
          sim = sims, 
          bins = "pretty",
          n_bins = 10,
          sim_cols = list(dv = 'CP',
                          idv = 'TIME',
                          id = 'ID',
                          sim = 'irep'
                          ),
          #log_y = TRUE,
          #stratify = 'DOSCOV',
          pi = c(0.025, 0.975),
          ci = c(0.025, 0.975),
          show = list(obs_dv = TRUE
                      #obs_ci = FALSE
                      #obs_median = FALSE
                      ),
          vpc_theme = mrg_vpc_theme,
          ylab  = 'Busuflan Concentration (mg/L)',
          xlab  = 'Time (hours)',
          title = "Prediction-Corrected Visual Predictive Check - pcVPC"
          ) + 
  theme_bw() +
  scale_x_continuous(limits = c(150, 600),
                     breaks = seq(0, 650, by = 30),
                     labels = function(x) x / 60)

vpc_plot +
  # Add dummy geoms to create legend entries
  geom_rect(aes(xmin = 0, xmax = 0, ymin = 0, ymax = 0, fill = "PI (95%)"), alpha = 0.5) +
  geom_rect(aes(xmin = 0, xmax = 0, ymin = 0, ymax = 0, fill = "Median"), alpha = 0.5) +
  
  # Define manual fill scale
  scale_fill_manual(name = "Simulated Data", 
                    values = c("PI (95%)" = "steelblue3", "Median" = "grey60")) +
  
  # Customize the legend position inside plot
  theme(
    legend.position = c(0.90, 0.80),   # x and y coordinates (0~1 scale)
    legend.background = element_rect(fill = "white", color = "black"),
    legend.title = element_text(size = 10),
    legend.text = element_text(size = 9)
  ) +
  
  guides(fill = guide_legend(override.aes = list(alpha = 0.5)))
```

## Estimating Individual exposure with TDM
```{r}
# Plasma drug concentration decline in two phases shown in semi-log plot
ggplot(busulfan_TDM_dataset, aes(TIME/60, DV, group = ID, color = as.factor(ID))) +
  geom_line() +
  geom_point() +
  theme_bw() +
  labs(title = "Busulfan First Dose semi-log plot", x = "Time (hours)", y = "Conc (mg/L)") +
  theme(legend.position = "blank")
```


## model fit
```{r}
busulfan_final_2cmt_model <- function() {
  ini({
    # Typical value (THETAs)
    tvcl   <- log(0.181)
    tvv1  <- log(25.34)
    tvq   <- log(0.355)
    tvv2  <- log(17.52)
    
    covbsav1 <- log(1.49)
    covbsacl <- log(1.03)
    covsexv2 <- log(0.757)
    
    # Interindividual variability (OMEGAs)
    eta_cl  ~ 0.0327
    eta_v1  ~ 0.0111
    eta_v2  ~ 0.0237
    
    # Residual variability
    prop.err <-  0.05313
  })
  
  model({
    # Individual value
    cl <- exp(tvcl + eta_cl) * (BSA/2.01)^covbsacl
    v1 <- exp(tvv1 + eta_v1) * (BSA/2.01)^covbsav1
    q  <- exp(tvq)
    v2 <- exp(tvv2 + eta_v2)
    
    # Sex effect
    if (Sex == 1) {
      v2 <- v2 * covsexv2
    } else {
      v2 <- v2 * 1
    }
    
    # Conversion
    k10 <- cl / v1
    k12 <- q / v1
    k21 <- q / v2
    
    # ODEs
    d / dt(central) = -k10 * central - k12 * central + k21 * peripheral
    d / dt(peripheral) = -k21 * peripheral + k12 * central
    
    # Concentration
    cp = central / v1
    
    # Error model (on log scale)
    IPRED = cp
    IPRED ~ prop(prop.err)
  })
}

# Model Fitting - Need MaxEval zero
two_cmt_TDM_pk_fit <- nlmixr2(
  busulfan_final_2cmt_model,
  busulfan_TDM_dataset,
  "posthoc",
  control = foceiControl(maxOuterIterations=0),
  table = list(cwres = TRUE, npde = TRUE)
)
```

## Estimating individual exposures with last time point
```{r}
last_time_point <- busulfan_TDM_dataset %>% 
  group_by(ID) %>%
  filter(row_number() == 1 | row_number() == 3)

# Plasma drug concentration decline in two phases shown in semi-log plot
ggplot(last_time_point, aes(TIME/60, DV, group = ID, color = as.factor(ID))) +
  geom_point() +
  theme_bw() +
  labs(title = "Busulfan First Dose semi-log plot", x = "Time (hours)", y = "Conc (mg/L)") +
  theme(legend.position = "blank")
```

## model fit
```{r}
busulfan_final_2cmt_model <- function() {
  ini({
    # Typical value (THETAs)
    tvcl   <- log(0.181)
    tvv1  <- log(25.34)
    tvq   <- log(0.355)
    tvv2  <- log(17.52)
    
    covbsav1 <- log(1.49)
    covbsacl <- log(1.03)
    covsexv2 <- log(0.757)
    
    # Interindividual variability (OMEGAs)
    eta_cl  ~ 0.0327
    eta_v1  ~ 0.0111
    eta_v2  ~ 0.0237
    
    # Residual variability
    prop.err <-  0.05313
  })
  
  model({
    # Individual value
    cl <- exp(tvcl + eta_cl) * (BSA/2.01)^covbsacl
    v1 <- exp(tvv1 + eta_v1) * (BSA/2.01)^covbsav1
    q  <- exp(tvq)
    v2 <- exp(tvv2 + eta_v2)
    
    # Sex effect
    if (Sex == 1) {
      v2 <- v2 * covsexv2
    } else {
      v2 <- v2 * 1
    }
    
    # Conversion
    k10 <- cl / v1
    k12 <- q / v1
    k21 <- q / v2
    
    # ODEs
    d / dt(central) = -k10 * central - k12 * central + k21 * peripheral
    d / dt(peripheral) = -k21 * peripheral + k12 * central
    
    # Concentration
    cp = central / v1
    
    # Error model (on log scale)
    IPRED = cp
    IPRED ~ prop(prop.err)
  })
}

# Model Fitting
two_cmt_last_tp_pk_fit <- nlmixr2(
  busulfan_final_2cmt_model,
  last_time_point,
  "posthoc",
  control = foceiControl(maxOuterIterations=0),
  table = list(cwres = TRUE, npde = TRUE)
)
```

## Estimating individual exposure with no observation
```{r}
## No observation
no_tp_AUC <- indv_fits %>%
  group_by(ID) %>%
  summarize(no_tp_AUC = first(no_obs_AUC), .groups = "drop")
```


## Full samples 
```{r}
# Plasma drug concentration decline in two phases shown in semi-log plot
ggplot(busulfan_First_Dose_dataset, aes(TIME/60, DV, group = ID, color = as.factor(ID))) +
  geom_point() +
  geom_line() +
  theme_bw() +
  labs(title = "Busulfan First Dose semi-log plot", x = "Time (hours)", y = "Conc (mg/L)") +
  theme(legend.position = "blank")
```

## model fit - full estimation
```{r}
busulfan_first_dose_2cmt_model <- function() {
  ini({
    # Typical value (THETAs)
    tvcl   <- log(0.187)
    tvv1  <- log(29)
    tvq   <- log(0.41)
    tvv2  <- log(17.3)
    
    covbsav1 <- log(2.32)
    covbsacl <- log(1.30)
    covsexv2 <- log(0.8)
    
    # Interindividual variability (OMEGAs)
    eta_cl  ~ 0.0222 
    eta_v1  ~ 0.0222
    eta_v2  ~ 0.0241
    
    # Residual variability
    prop.err <-  0.0955
  })
  
  model({
    # Individual value
    cl <- exp(tvcl + eta_cl) * (BSA/2.01)^covbsacl
    v1 <- exp(tvv1 + eta_v1) * (BSA/2.01)^covbsav1
    q  <- exp(tvq)
    v2 <- exp(tvv2 + eta_v2)
    
    # Sex effect
    if (Sex == 1) {
      v2 <- v2 * covsexv2
    } else {
      v2 <- v2 * 1
    }
    
    # Conversion
    k10 <- cl / v1
    k12 <- q / v1
    k21 <- q / v2
    
    # ODEs
    d / dt(central) = -k10 * central - k12 * central + k21 * peripheral
    d / dt(peripheral) = -k21 * peripheral + k12 * central
    
    # Concentration
    cp = central / v1
    
    # Error model (on log scale)
    IPRED = cp
    IPRED ~ prop(prop.err)
  })
}

# Model Fitting
full_estimation_pk_fit <- nlmixr2(
  busulfan_first_dose_2cmt_model,
  busulfan_First_Dose_dataset,
  "focei",
  table = list(cwres = TRUE, npde = TRUE)
)
```

## Estimating individual Exposures - Full estimation, TDM, Last Timepoint, No observation
```{r}
## TDM
orig_data <- two_cmt_TDM_pk_fit$origData
indv_cl <- data.frame(ID = two_cmt_TDM_pk_fit$ID, Clearance = two_cmt_TDM_pk_fit$cl)

tdm_indiv_AUC<- merge(orig_data, indv_cl, by = 'ID')
tdm_indiv_AUC <- tdm_indiv_AUC %>% mutate(AUC = AMT/Clearance) %>% select(ID, AUC)

tdm_indiv_AUC <- tdm_indiv_AUC %>%
  group_by(ID) %>%
  summarize(tdm_AUC = first(AUC), .groups = "drop")

## Last Time POint
last_tp_indv_cl <- data.frame(ID = two_cmt_last_tp_pk_fit$ID, Clearance = two_cmt_last_tp_pk_fit$cl)

last_tp_indiv_exposure <- merge(last_time_point, last_tp_indv_cl, by = 'ID')
last_tp_indiv_exposure <- last_tp_indiv_exposure %>% mutate(AUC = AMT/Clearance) %>% select(ID, AUC)

last_tp_AUC <- last_tp_indiv_exposure %>%
  group_by(ID) %>%
  summarize(last_tp_AUC = first(AUC), .groups = "drop")

## Full-Estimation
full_est_indv_cl <- data.frame(ID = full_estimation_pk_fit$ID, Clearance = full_estimation_pk_fit$cl)

full_est_indiv_exposure <- merge(busulfan_First_Dose_dataset, full_est_indv_cl, by = 'ID')
full_est_indiv_exposure <- full_est_indiv_exposure %>% mutate(AUC = AMT/Clearance) %>% select(ID, AUC)

full_est_AUC <- full_est_indiv_exposure %>%
  group_by(ID) %>%
  summarize(full_estimation_AUC = first(AUC), .groups = "drop")

compare_auc <- merge(full_est_AUC, tdm_indiv_AUC, by = "ID")
compare_auc <- merge(compare_auc, last_tp_AUC, by = "ID")
compare_auc <- merge(compare_auc, no_tp_AUC, by = "ID")

compare_auc[] <- lapply(compare_auc, function(x) if(is.numeric(x)) round(x, 0) else x)
print(compare_auc)
```

## Compare across estimation methods
```{r}
# Convert to long format
df_long <- compare_auc %>%
  pivot_longer(cols = -ID, names_to = "Method", values_to = "AUC")

# Order methods for nicer plotting
df_long$Method <- factor(df_long$Method, levels = c("full_estimation_AUC", "tdm_AUC", "last_tp_AUC", "no_tp_AUC"))

# Set common y-axis limits
y_limits <- c(750, 1750)

# --- Full vs TDM
df_tdm <- compare_auc %>%
  select(ID, Full = full_estimation_AUC, TDM = tdm_AUC) %>%
  pivot_longer(cols = -ID, names_to = "Method", values_to = "AUC") %>%
  mutate(Method = factor(Method, levels = c("Full", "TDM")))  

p_tdm <- ggplot(df_tdm, aes(x = Method, y = AUC)) +
  geom_boxplot(alpha = 0.3, aes(fill = Method, color = Method)) +
  geom_jitter(aes(color = Method), width = 0.1, alpha = 0.7, size = 2) +
  geom_line(aes(group = ID), color = "grey60", alpha = 0.6) +
  scale_fill_manual(values = c("Full" = "#F8766D", "TDM" = "#00BFC4")) +
  scale_color_manual(values = c("Full" = "#F8766D", "TDM" = "#00BFC4")) +
  theme_bw() +
  labs(title = "TDM", x = "", y = "AUC") +
  theme(legend.position = "none",
        plot.title = element_text(hjust = 0.5)
        ) +
  scale_y_continuous(limits = y_limits)

# --- Full vs Last TP
df_last <- compare_auc %>%
  select(ID, Full = full_estimation_AUC, Last_TP = last_tp_AUC) %>%
  pivot_longer(cols = -ID, names_to = "Method", values_to = "AUC") %>%
  mutate(Method = factor(Method, levels = c("Full", "Last_TP")))  

p_last <- ggplot(df_last, aes(x = Method, y = AUC)) +
  geom_boxplot(alpha = 0.3, aes(fill = Method, color = Method)) +
  geom_jitter(aes(color = Method), width = 0.1, alpha = 0.7, size = 2) +
  geom_line(aes(group = ID), color = "grey60", alpha = 0.6) +
  scale_fill_manual(values = c("Full" = "#F8766D", "Last_TP" = "#7CAE00")) +
  scale_color_manual(values = c("Full" = "#F8766D", "Last_TP" = "#7CAE00")) +
  theme_bw() +
  labs(title = "Last TP", x = "", y = "AUC") +
  theme(legend.position = "none",
        axis.title.y = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        plot.title = element_text(hjust = 0.5)
        ) +
  scale_y_continuous(limits = y_limits)

# --- Full vs No TP
df_no <- compare_auc %>%
  select(ID, Full = full_estimation_AUC, No_Obs = no_tp_AUC) %>%
  pivot_longer(cols = -ID, names_to = "Method", values_to = "AUC") %>%
  mutate(Method = factor(Method, levels = c("Full", "No_Obs")))  

p_no <- ggplot(df_no, aes(x = Method, y = AUC)) +
  geom_boxplot(alpha = 0.3, aes(fill = Method, color = Method)) +
  geom_jitter(aes(color = Method), width = 0.1, alpha = 0.7, size = 2) +
  geom_line(aes(group = ID), color = "grey60", alpha = 0.6) +
  scale_fill_manual(values = c("Full" = "#F8766D", "No_Obs" = "#C77CFF")) +
  scale_color_manual(values = c("Full" = "#F8766D", "No_Obs" = "#C77CFF")) +
  theme_bw() +
  labs(title = "No Obs", x = "", y = "AUC") +
  theme(legend.position = "none",
        axis.title.y = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        plot.title = element_text(hjust = 0.5)
        ) +
  scale_y_continuous(limits = y_limits)

wrap_plots(p_tdm, p_last, p_no, ncol = 3)
```


