---
title: "2025 National Taiwan University - Population Pharmacokinetics workshop: TEST"
author: "Jin Gyu Kim"
date: "2025-05-02"
output:
  html_document:
    css: docs/src/styles/styles.css
    toc: true
    number_sections: true
---

***
1. **Install the C Compiler**  
Ensure that a C compiler is installed on your system. For installation instructions, refer to `ReadMe.txt`.

2. **Activate the R environment**  
Run the following code chunk to install *renv* and *pkgbuild*. When prompted in the console, type Y at install.packages()
```{r}
install.packages("renv")
```

After installing packages, activate the R project by running the code chunk below
```{r}
library(renv)
renv::activate()
```

2-1. **Restore packages**
This step will allow you to sync the packages from renv.lock file.
When prompted in the console, type Y to update your packages inside the project folder.
```{r, echo = TRUE, message = "hide", warning = "hide", results = "hide"}
renv::restore()
```

3. **Run Script**
Run the scripts below and check if it runs successfully

***

**Load Packages**
```{r echo = TRUE, message = "hide", warning = "hide", results = "hide"}
rm(list=ls())
library(tidyverse)
library(ggplot2)
library(nlmixr2)
library(xpose4)
library(xpose.nlmixr2)
library(rxode2)
library(gridExtra)
library(ggPMX)
library(ggpubr)
library(mrgsolve)
```

**Data Import**
```{r, warning = "hide"}
# Import busulfan two compartment dataset
busulfan_dataset <- read.csv("dataset/busulfan_First_Dose.csv", na.strings = ".")

# Plasma drug concentration decline in two phases shown in semi-log plot
ggplot(busulfan_dataset, aes(TIME/60, log(DV), group = ID, color = as.factor(ID))) +
  geom_line() +
  theme_bw() +
  labs(title = "Busulfan semi-log plot", x = "Time (hours)", y = "Conc (mg/L)") +
  theme(legend.position = "blank")
```

**Fit**
```{r, results = "hide"}
# One compartment model structure
busulfan_1cmt_base_model <- function() {
  ini({
    # Typical value (THETAs)
    tvcl   <- log(5)
    tvv1  <- log(50)

    # Interindividual variability (OMEGAs)
    eta_cl  ~ 0.0322
    eta_v1  ~ 0.0222

    # Residual variability
    add.error <-  0.0955
  })

  model({
    # Individual value
    cl <- exp(tvcl + eta_cl)
    v1 <- exp(tvv1 + eta_v1)

    # Conversion
    k10 <- cl / v1

    # ODEs
    d / dt(central) = -k10 * central

    # Concentration
    cp = central / v1

    # Error model
    IPRED = cp
    IPRED ~ add(add.error)
  })
}

# 1 compartment model fitting
one_cmt_pk_fit <- nlmixr2(
  busulfan_1cmt_base_model,
  busulfan_dataset,
  "focei",
  table = list(cwres = TRUE)
)
```

**Plot**
```{r, warning = "hide"}
# Convert to XPOSE for goodness of fit
one_cmt_xpdb_pk <- xpose_data_nlmixr2(one_cmt_pk_fit,
                                     xp_theme = theme_xp_nlmixr2()
                                     )

# observation vs population prediction plot
dv_vs_pred(one_cmt_xpdb_pk,
           type="pls",
           title = "One Compartment",
           log   = 'xy',
           subtitle = NULL,
           guide = TRUE)
```
